---
title: "Introduction to finclean package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to finclean package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(finclean)
```

## What is finclean?

`finclean` is an R package that helps you clean and diagnose messy financial
data. Real-world financial datasets often have inconsistent formats, missing
values, and suspicious entries that need to be addressed before any analysis.
This vignette walks through each function using the built-in sample dataset.

## Load the Dataset

`finclean` comes with a built-in sample dataset `finclean_sample` that contains
common data quality issues found in financial data — messy currency strings,
inconsistent fiscal year formats, account name variations, outliers, and missing
values.

```{r}
data(finclean_sample)
finclean_sample
```

## Parse Currency Strings

Financial data often contains currency values stored as strings with symbols,
commas, and parentheses. `parse_currency()` converts these into clean numeric
values. Parentheses-style negatives like `"(1,200)"` are correctly converted
to `-1200`.

```{r}
parse_currency(finclean_sample$amount)
```

## Standardize Fiscal Year Formats

Fiscal year labels are often inconsistent across teams and systems.
`standardize_fiscal_year()` normalizes all formats into a consistent
`FY####` or `FY####-Q#` output.

```{r}
standardize_fiscal_year(finclean_sample$period)
```

## Normalize Account Names

Account names are frequently abbreviated or capitalized differently by
different people. `normalize_accounts()` maps all variations to a single
standardized label. Note that if you supply a `custom_map`, it will
completely replace the built-in mapping.

```{r}
normalize_accounts(finclean_sample$account)
```

## Flag Outliers

`flag_outliers()` detects suspicious values in a numeric column using either
the IQR method (default) or the z-score method. It returns a data frame with
each value labeled as an outlier or not.

```{r}
flag_outliers(finclean_sample$revenue)
```

## Audit the Full Dataset

Instead of checking each column individually, `audit_report()` scans the
entire data frame at once and returns a structured summary of missing values,
duplicate counts, and outliers for every column.

```{r}
audit_report(finclean_sample)
```

## Visualize Outliers

`plot_outliers()` generates a bar chart that highlights outlier observations
in red, making it easy to spot anomalies at a glance.

```{r, fig.width=6, fig.height=4}
plot_outliers(finclean_sample, column = "revenue")
```
